Formal Verification Framework Based on Symbolic Execution for Smart Contract

### 2023-12-14 papers/static-checker/ecoop2024/ecoop2024.tex
write static checker part

%FAILWITH
\begin{mathpar}
  \inferrule[FAILWITH]
  {
  }{[(\FAILWITH; \INSTRUCTION), (\StackOne,  \TY) \STACKCONCAT \STACK,  \PREDICATE] \StateTrans [\EMPTYSTACK, (\FAIL\ (\StackOne), \TFAILWITH) \STACKCONCAT\EMPTYSTACK, \PREDICATE]}
\end{mathpar}
\section {Domain Specific Language For Smart Contract Property Specification}
Michelson, as a stack-based language, may pose a challenge for developers less familiar with it compared to more mainstream languages like Python, Java, or C. To address this, several high-level languages such as LIGO, SmartPy, and Liquidity have been created. These languages provide a more readable syntax, resembling popular programming languages, allowing developers to write Tezos smart contracts more comfortably. These high-level contracts are then compiled down to Michelson.

@@

\begin{lstlisting}[float,captionpos=b,caption={Smart contract property specification},label={lst:smart-contract-property-specification},numbers=left]
mcontract Name = spec
code := {c}
input := t
output := t'
pre-condition := cd
post-condition := cd'
\end{lstlisting}
Listing \ref{lst:smart-contract-property-specification} shows the syntax of our property specification language. Here, \KMCONTRACT, \KWITH, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are keywords. The \KNAME\ is replaced by the contract name. In the input and output fields, the input and output patterns are specified. Since Michelson operates on a stack, we assume that the stack has only one element, as specified in the input pattern. Similarly, the output stack is assumed to be a stack with only one element, which is specified in the output field. As we have mentioned before, the output should be a pair consisting of an operation list and a storage. In our syntax, for verification purpose, we omit the operation list and then the output is represented by the patern of the value in the storage after the execution is finished. The input and output patterns represent Michelson terms, which can be either concrete values or symbolic values, including variables. When a variable appears for the first time, it needs to be accompanied by its corresponding type declaration. However, subsequent mentions can omit the type declaration. Formally, the pre and post-condition instances are formulas in first-order logic, and they can be handled by an SMT solver, namely the Z3 solver in our framework. The detail of the language grammar is described in in Figure \ref{}.

@@

\section {Static Checker}
The architect of the static checker is depicted as in Figure \ref{}.  The system includes

Formulas that are used in assertions or added to solvers are terms of Boolean sort. Otherwise, terms of Boolean and non-Boolean sort may be mixed in any combination where sorts match up. Universal and existential quantifiers bind variables to the scope of the quantified formula.
\section {Case Studies}
\subsection{USDtz}
\subsection{Kolibri Oracle Contract}
\section {Related Work}
\section {Conclusion}
\section {Appendix}
\end{document}

>>>>

%FAILWITH
\begin{mathpar}
  \inferrule[FAILWITH]
  {
  }{[(\FAILWITH; \INSTRUCTION), (\StackOne,  \TY) \STACKCONCAT \STACK,  \PREDICATE] \StateTrans [\EMPTYSTACK, (\FAIL\ (\StackOne), \TFAILWITH) \STACKCONCAT\EMPTYSTACK, \PREDICATE]}
\end{mathpar}
\section {Domain Specific Language For Smart Contract Property Specification}
\label{sec:domain-specific-language}
Michelson, as a stack-based language, may pose a challenge for developers less familiar with it compared to more mainstream languages like Python, Java, or C. To address this, several high-level languages such as LIGO, SmartPy, and Liquidity have been created. These languages provide a more readable syntax, resembling popular programming languages, allowing developers to write Tezos smart contracts more comfortably. These high-level contracts are then compiled down to Michelson.

@@

\begin{lstlisting}[float,captionpos=b,caption={Smart contract property specification},label={lst:smart-contract-property-specification},numbers=left]
mcontract Name = spec
code := {c}
input := t
output := t'
pre-condition := cd
post-condition := cd'
\end{lstlisting}
Listing \ref{lst:smart-contract-property-specification} shows the syntax of our property specification language. Here, \KMCONTRACT, \KWITH, \KCODE, \KINPUT, \KOUTPUT, \KPRECONDI, and \KPOSTCONDI\ are keywords. The \KNAME\ is replaced by the contract name. In the input and output fields, the input and output patterns are specified. Since Michelson operates on a stack, we assume that the stack has only one element, as specified in the input pattern. Similarly, the output stack is assumed to be a stack with only one element, which is specified in the output field. As we have mentioned before, the output should be a pair consisting of an operation list and a storage. In our syntax, for verification purpose, we omit the operation list and then the output is represented by the patern of the value in the storage after the execution is finished. The input and output patterns represent Michelson terms, which can be either concrete values or symbolic values, including variables. When a variable appears for the first time, it needs to be accompanied by its corresponding type declaration. However, subsequent mentions can omit the type declaration. Formally, the pre and post-condition instances are formulas in first-order logic, and they can be handled by an SMT solver, namely the Z3 solver in our framework. The detail of the language grammar is described in in Figure \ref{}. Formulas that are used in assertions or added to solvers are terms of Boolean sort. Otherwise, terms of Boolean and non-Boolean sort may be mixed in any combination where sorts match up. Universal and existential quantifiers bind variables to the scope of the quantified formula.

@@

\section {Static Checker}
We verify the contract through symbolic execution against its specifications in the contract module. Symbolic execution involves calculating a symbolic stack at each transition from one Michelson instruction to the next. The symbolic interpreter is fully typed and rejects ill-typed Michelson programs, provided the ill-typed part is reachable from the default entry point. The initial stack is derived from the input pattern for each the entry point. To maintain symbolic values as concrete as possible, some instructions may be deemed unreachable or only reachable under specific conditions. This process yields, for each entry point, a set of final symbolic stacks accompanied by a path condition indicating when this final state is reachable. A state is discarded from the final state space if the path condition is not satisfied, indicating that the path is unreachable. Additionally, for each \FAILWITH\ instruction, we obtain a symbolic value for the reported message and a path condition indicating the reachability of this instruction. We employ the SMT solver Z3  to assess the feasibility of a path condition and to check smart contract properies specifying by pre and post conditions.

The architecture of the static checker is illustrated in Figure \ref{}. The system comprises three phases. The first phase involves user specification, where users provide the Michelson code of the contract and define the properties that the smart contract should satisfy. Subsequently, the code, along with the input pattern, is input to the interpreter to achieve symbolic execution. The path conditions are then fed into a Z3 converter, which transforms these symbolic terms into Z3 formulas. Concurrently, the property specification, specifically pre and post conditions, also undergoes conversion to Z3 formulas in the intermediate phase. The results of the symbolic execution and Z3 formulas are then input into the static checker, which verifies whether these properties hold for the smart contract implementation. This constitutes the verification phase.

\subsection{User specification}
In the user specification phase, users compose the specification in our domain-specific language using the syntax outlined in Section \ref{sec:domain-specific-language}. Note that Michelson code, whether authored by users, obtained from the Tezos blockchain, or compiled from higher-level languages, can be seamlessly employed within our framework. The specification is subsequently parsed into an Abstract Syntax Tree (AST) in OCaml. The code and the input pattern are then utilized as inputs for the interpreter during the intermediate phase to facilitate symbolic execution. In the next phase, the pre and post predicates are transformed into Z3 formulas by the Z3 converter tool.
\subsection{Michelson interpreter and Z3 converter}
The interpreter implements the execution model described in \ref{sec:symbolic-execution-model}. When provided with a concrete input, the interpreter produces a concrete value. In the presence of symbolic input, the result comprises a set of symbolic states represented as terms in OCaml. Due to Michelson being built in OCaml, the conversion of Michelson types or terms to their OCaml counterparts is straightforward. Each state contains the output stack, which has only one element consisting of an operation list and storage, along with the path conditions represented as predicates in conjunctive form. This output is then fed into the Z3 converter, which handles two different types of terms: assertion terms from the AST representing the pre and post-condition specifications, and terms resulting from the interpreter. We utilize the Z3 library in OCaml to construct this converter. Some adjustments are necessary to convert these terms into Z3 formulas. For example, as there is no natural number type (\TNAT) in Z3, the term is converted as an integer type, accompanied by constraints asserting that it is greater than or equal to zero. Several constants, such as \CAMOUNT, \CBALANCE, \CSENDER\ and cryptographic functions, such as \FSHA, \FHASHKEY\ are predefined in the environment that works for any smart contract specifications.
\subsection{Static checker}
The static checker now performs the verification process by employing symbolic execution against the property specification in the DSL. It utilizes the results from the previous phases, including symbolic execution, Z3 formulas representing pre and post conditions, and path conditions. The static checker performs three distinct features: the output check, fail condition and property check.

\subsubsection{Output check} First is the output check, where the checker verifies whether the output pattern matches. It basiclly matches the output pattern with the correcspoding part in the final states resulting from the symbolic intepreter.  When the input pattern is a concrete value, the checker can be used as a test to verify if the output aligns with the expectations specified in the output field.

\subsubsection{Fail condition}

\subsubsection{Property check}

consider the code is as a black box
\section {Case Studies}
\subsection{USDtz}
\subsection{Kolibri Oracle Contract}
\section {Related Work}
\section {Conclusion}
\section {Appendix}
\end{document}
